---
title: "Comparing clades: Caligus vs Lepeophtheirus sea lice on marine fishes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing clades: Caligus vs Lepeophtheirus sea lice on marine fishes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_cofid <- requireNamespace("cofid", quietly = TRUE)
```

This vignette shows the full phylopred v2 workflow on a real host-parasite
system: sea lice (Copepoda: Caligidae) parasitizing marine fishes. We ask
whether the probability that two fish species share a sea louse decays
faster with host phylogenetic distance in *Lepeophtheirus* than in
*Caligus* — i.e., whether *Lepeophtheirus* is a phylogenetic specialist
and *Caligus* a generalist.

Interaction records come from the
[cofid](https://github.com/alrobles/cofid) package (copepod-fish
interaction database) and the host phylogeny from the
[Fish Tree of Life](https://fishtreeoflife.org/) (Rabosky et al. 2018),
pruned to the recorded hosts and shipped as the `fish_tree` data object.

```{r setup, eval = has_cofid}
library(phylopred)
library(ape)
```

## 1. Interaction data

We keep one record per copepod species / fish species pair and derive
the copepod genus, which is the clade-level grouping we want to compare.

```{r data, eval = has_cofid}
interactions <- cofid::cofid
interactions$genus <- sub(" .*", "", interactions$source_taxon_name)
interactions <- interactions[
  interactions$genus %in% c("Caligus", "Lepeophtheirus"),
  c("source_taxon_name", "target_taxon_name", "genus")
]
table(unique(interactions[, c("source_taxon_name", "genus")])$genus)
```

## 2. Host phylogenetic distances

`fish_tree` is already pruned to the recorded hosts present in the Fish
Tree of Life; its cophenetic matrix gives the phylogenetic distance
between every pair of hosts.

```{r phydist, eval = has_cofid}
data(fish_tree)
phydist <- cophenetic(fish_tree)
dim(phydist)
```

## 3. Pairwise host-sharing data

`prepare_pair_data()` expands the interaction table into all
(focal host, target host) pairs per parasite. Every known host is used
as a focal host — unlike the legacy workflow, which sampled a single
focal host per parasite — so the data set is deterministic and uses all
available information.

```{r pairs, eval = has_cofid}
pairs <- prepare_pair_data(interactions, phydist, group = "genus")
nrow(pairs)
head(pairs)
```

## 4. Comparing slopes between clades

`compare_phylopred_slopes()` fits a single logistic regression with a
genus-by-distance interaction and quantifies uncertainty with a
parasite-level cluster bootstrap, which respects the non-independence
of pairs that share a parasite.

```{r compare, eval = has_cofid}
cmp <- compare_phylopred_slopes(pairs, n_boot = 20, seed = 42)
cmp
```

A more negative slope means the probability of sharing a parasite
decays faster with phylogenetic distance. The *Lepeophtheirus* slope is
steeper than the *Caligus* slope and the interaction term is negative,
supporting the interpretation of *Lepeophtheirus* as a phylogenetic
specialist and *Caligus* as a generalist.

## 5. Predicted sharing curves

```{r curves, eval = has_cofid}
cf <- coef(cmp$fitted_model)
d <- seq(0, max(pairs$phydist), length.out = 200)
p_cal <- plogis(cf["(Intercept)"] + cf["phydist"] * d)
p_lep <- plogis(cf["(Intercept)"] + cf["genusLepeophtheirus"] +
                  (cf["phydist"] + cf["phydist:genusLepeophtheirus"]) * d)
plot(d, p_cal, type = "l", col = "#1b9e77", lwd = 2, ylim = c(0, max(p_cal)),
     xlab = "Host phylogenetic distance",
     ylab = "P(sharing a sea louse)")
lines(d, p_lep, col = "#d95f02", lwd = 2)
legend("topright", legend = c("Caligus (generalist)",
                              "Lepeophtheirus (specialist)"),
       col = c("#1b9e77", "#d95f02"), lwd = 2, bty = "n")
```

## 6. Model capacity and joint uncertainty

`evaluate_phylopred_model()` reports pseudo-R² and AUC (optionally with a
parasite-grouped cross-validation, which measures the ability to predict
the host range of unseen parasites), and
`phylopred_confidence_ellipse()` gives the joint confidence region of
intercept and slope from the cluster bootstrap.

```{r evaluate, eval = has_cofid}
evaluate_phylopred_model(pairs)
```

```{r ellipse, eval = has_cofid}
ell <- phylopred_confidence_ellipse(cmp$bootstrap)
plot(ell, type = "l",
     main = "95% joint confidence ellipse (cluster bootstrap)")
points(t(attr(ell, "center")), pch = 19)
```

## 7. Presence-only threshold metrics: f_gamma and partial ROC

Interaction data are presence-only: unrecorded host-parasite pairs are
unlabeled, not true absences. `host_threshold_metric()` adapts the
SDM threshold metric \(f_\gamma(t) = a/(\gamma b + c)\) to phylogenies —
`a` is the sensitivity over known hosts and `b` the fraction of all
species in the phylogeny predicted as hosts — and
`phylo_partial_roc()` gives the Peterson-Papes-Soberon partial AUC
ratio against a null (random-ranking) model.

```{r fgamma, eval = has_cofid}
fit <- glm(suscept ~ phydist, data = pairs, family = binomial())
example_parasite <- names(sort(table(pairs$parasite), decreasing = TRUE))[1]
one <- pairs[pairs$parasite == example_parasite, ]
one$pred <- predict(fit, newdata = one, type = "response")
# Suitability of each target species: best score across the parasite's
# known focal hosts
agg <- aggregate(cbind(pred, suscept) ~ target, data = one, FUN = max)

m <- host_threshold_metric(agg$pred, agg$suscept, gamma = 1)
m
plot(m)
```

With `gamma = 1`, `f_gamma` can be read as a phylogenetic efficiency:
known hosts retained per unit of phylogeny predicted as suitable.

In Peterson, Papes & Soberon (2008) the tolerance `E` bounds the
omission error on the y axis, motivated by the known georeferencing /
identification error of occurrence databases. For interaction records we
do not know that error, so `E` is simply the omission we are willing to
tolerate; a large value (here `E = 0.95`) uses essentially the whole
curve, so the ratio compares the full presence-only AUC against the
random-ranking null.

```{r partialroc, eval = has_cofid}
roc <- phylo_partial_roc(agg$pred, agg$suscept, error_rate = 0.95,
                         n_boot = 200, seed = 42)
roc
plot(roc)
```

## 8. Positive-unlabeled prediction with xplus

The pairwise GLM scores every candidate host through a parametric
curve in phylogenetic distance. `fit_phylopred_pu()` is a parallel,
non-parametric route: known hosts of one parasite are the positives,
every other tip of the phylogeny is unlabeled (not a confirmed
absence), and an iterative PU learner (`xplus`) is trained on
phylogenetic features (minimum and mean distance to the known hosts
plus PCoA axes of the tree).

```{r pu, eval = has_cofid}
pu <- fit_phylopred_pu(interactions, phydist, example_parasite,
                    seed = 42, max_iter = 30)
pu
m_pu <- host_threshold_metric(pu$prediction$pred, pu$prediction$is_host)
roc_pu <- phylo_partial_roc(pu$prediction$pred, pu$prediction$is_host,
                            error_rate = 0.95, n_boot = 200, seed = 42)
roc_pu
```

Because the unlabeled species are not true absences, in-sample
classification metrics overstate performance. `validate_host_prediction()`
hides a fraction of the known hosts, refits, and measures how well the
model ranks the hidden hosts above truly unlabeled species:

```{r holdout, eval = has_cofid}
validate_host_prediction(interactions, phydist, example_parasite,
                         method = "glm", n_rep = 5, seed = 42)
validate_host_prediction(interactions, phydist, example_parasite,
                         method = "pu", n_rep = 5, seed = 42, max_iter = 20)
```

This holdout AUC is a ranking/recovery score against unlabeled species,
not a classical ROC against confirmed negatives.

## References

- Robles-Fernández, Á. L. & Lira-Noriega, A. (2017). Combining
  phylogenetic and occurrence information for risk assessment of pest
  and pathogen interactions with host plants. *Frontiers in Applied
  Mathematics and Statistics*, 3:17.
  [doi:10.3389/fams.2017.00017](https://doi.org/10.3389/fams.2017.00017)
- Rabosky, D. L. et al. (2018). An inverse latitudinal gradient in
  speciation rate for marine fishes. *Nature*, 559, 392-395.
- Tinoco-Domínguez, E., Amancio, G., Robles-Fernández, Á. L. &
  Lira-Noriega, A. (2025). Interaction network of *Phoradendron* and its
  hosts. *American Journal of Botany*, e70025.
- cofid: curated copepod-fish interaction database.
  https://github.com/alrobles/cofid
