---
title: "Specialist vs generalist copepods: Caligus and Lepeophtheirus"
output: rmarkdown::html_document
bibliography: ../../paper/refs.bib
vignette: >
  %\VignetteIndexEntry{Specialist vs generalist copepods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  cache = TRUE
)
set.seed(42)
```

This article walks through the full `phylopred` workflow on a real
use case: contrasting a *generalist* sea-louse genus (*Caligus*)
against a more *specialist* one (*Lepeophtheirus*), using curated
copepod–fish interactions from the
[`cofid`](https://github.com/alrobles/cofid) database and host
phylogenetic distances from the Fish Tree of Life
(Rabosky et al. 2018; distributed with the package as `fish_tree`,
CC0). In the multi-genus comparative sweep these two genera sit at
opposite ends of the phylogenetic host-range gradient
(median $b^*$ = 0.266 for *Caligus* vs 0.140 for *Lepeophtheirus*).

Everything below is presence-only: recorded interactions are
positives, and fish without a record for a given parasite are treated
as *unlabeled*, never as confirmed absences.

```{r setup, message = FALSE}
library(phylopred)
library(cofid)
library(ape)
```

## 1. Build the interaction table and the candidate universe

We take all *Caligus* and *Lepeophtheirus* records from `cofid`,
keep hosts present in `fish_tree`, and compute patristic distances.
The pruned tree defines the *candidate host universe*: the set of
species over which suitability is predicted and coverage is measured.

```{r data}
data(cofid, package = "cofid")

genus <- sub(" .*", "", cofid$source_taxon_name)
keep <- genus %in% c("Caligus", "Lepeophtheirus")

interactions <- unique(data.frame(
  parasite = cofid$source_taxon_name[keep],
  host     = cofid$target_taxon_name[keep],
  genus    = genus[keep],
  stringsAsFactors = FALSE
))

tips <- gsub("_", " ", fish_tree$tip.label)
interactions <- interactions[interactions$host %in% tips, ]

phydist <- cophenetic(fish_tree)
rownames(phydist) <- colnames(phydist) <- gsub("_", " ", rownames(phydist))

nrow(interactions)
table(interactions$genus)
```

## 2. Probability of sharing vs phylogenetic distance

`prepare_pair_data()` expands, for each parasite, all pairs between
its known hosts and the candidate universe. A single logistic model
with a genus × distance interaction tests whether the two genera
decay differently — the formal replacement for comparing two
separately fitted regressions.

```{r slopes}
pairs <- prepare_pair_data(interactions, phydist, group = "genus")
cmp <- compare_phylopred_slopes(pairs, n_boot = 200, seed = 42)
cmp
```

A more negative slope means the probability of sharing a parasite
falls faster with phylogenetic distance — the specialist signature.
The cluster bootstrap resamples whole parasites, honestly propagating
the between-parasite heterogeneity into the intervals.

```{r curves, fig.cap = "Probability-of-sharing decay per genus."}
fit <- cmp$fitted_model
d <- seq(0, max(pairs$phydist), length.out = 200)
newdat <- rbind(
  data.frame(phydist = d, group = "Caligus"),
  data.frame(phydist = d, group = "Lepeophtheirus")
)
newdat$p <- predict(fit, newdata = newdat, type = "response")

plot(NULL, xlim = range(d), ylim = range(newdat$p),
     xlab = "Phylogenetic distance", ylab = "P(share parasite)")
lines(d, newdat$p[newdat$group == "Caligus"], col = "firebrick", lwd = 2)
lines(d, newdat$p[newdat$group == "Lepeophtheirus"], col = "steelblue", lwd = 2)
legend("topright", legend = c("Caligus (generalist)",
                              "Lepeophtheirus (specialist)"),
       col = c("firebrick", "steelblue"), lwd = 2, bty = "n")
```

## 3. Positive–unlabeled host prediction

For one well-sampled parasite of each genus we fit the PU classifier
(`xplus` engine) on phylogenetic features: minimum and mean distance
to the known hosts plus PCoA axes of the host distance matrix.

```{r pick}
host_counts <- sort(table(interactions$parasite), decreasing = TRUE)
head(host_counts)

caligus_sp <- "Caligus elongatus"
lepeo_sp   <- grep("^Lepeophtheirus", names(host_counts), value = TRUE)[1]
lepeo_sp
```

```{r pu, message = FALSE}
pu_cal <- fit_phylopred_pu(interactions, phydist, parasite = caligus_sp,
                           seed = 42)
pu_lep <- fit_phylopred_pu(interactions, phydist, parasite = lepeo_sp,
                           seed = 42)
pu_cal
head(pu_cal$prediction[order(-pu_cal$prediction$pred), ])
```

## 4. Presence-only performance and phylogenetic host range

`host_threshold_metric()` computes, over all thresholds $t$, the
efficiency $f_\gamma(t) = a/(\gamma b + c)$ where $a$ is the fraction
of known hosts retained, $b$ the fraction of the candidate phylogeny
predicted suitable, and $c = 1 - a$. The coverage $b^*$ at the
optimum is the *phylogenetic host range* — our generalism measure.

```{r fgamma}
m_cal <- host_threshold_metric(pu_cal$prediction$pred, pu_cal$prediction$is_host)
m_lep <- host_threshold_metric(pu_lep$prediction$pred, pu_lep$prediction$is_host)

summarize_fgamma <- function(m, parasite) {
  best <- which.max(m$f_gamma)
  data.frame(parasite = parasite,
             f_star = attr(m, "optimal_f"),
             t_star = attr(m, "optimal_threshold"),
             host_range = m$b[best],
             sensitivity = m$a[best])
}
rbind(summarize_fgamma(m_cal, caligus_sp),
      summarize_fgamma(m_lep, lepeo_sp))
```

```{r fgamma-plot, fig.cap = "f_gamma efficiency curves."}
op <- par(mfrow = c(1, 2))
plot(m_cal$threshold, m_cal$f_gamma, type = "s",
     xlab = "threshold t", ylab = expression(f[gamma]), main = caligus_sp)
abline(v = attr(m_cal, "optimal_threshold"), lty = 2)
plot(m_lep$threshold, m_lep$f_gamma, type = "s",
     xlab = "threshold t", ylab = expression(f[gamma]), main = lepeo_sp)
abline(v = attr(m_lep, "optimal_threshold"), lty = 2)
par(op)
```

The phylogenetic partial ROC compares the sensitivity–coverage curve
against the null diagonal $a = b$ (a random ranking of species). We
use `error_rate = 0.95` — since the database error rate is unknown,
$E$ is the omission we are willing to tolerate, and 0.95 uses
essentially the full curve.

```{r proc, fig.cap = "Phylogenetic partial ROC for the two parasites."}
roc_cal <- phylo_partial_roc(pu_cal$prediction$pred, pu_cal$prediction$is_host,
                             error_rate = 0.95, n_boot = 200, seed = 42)
roc_lep <- phylo_partial_roc(pu_lep$prediction$pred, pu_lep$prediction$is_host,
                             error_rate = 0.95, n_boot = 200, seed = 42)

op <- par(mfrow = c(1, 2))
plot(roc_cal)
mtext(caligus_sp, side = 3, line = -1, cex = 0.8)
plot(roc_lep)
mtext(lepeo_sp, side = 3, line = -1, cex = 0.8)
par(op)

c(caligus = roc_cal$auc_ratio, lepeophtheirus = roc_lep$auc_ratio)
```

AUC ratios above 1 mean known hosts are ranked above unlabeled
species more often than chance. This is a ranking statement — not
discrimination against true absences, which we do not have.

## 5. Out-of-sample validation: hiding known hosts

The decisive test of predictive value hides 25% of each parasite's
known hosts, refits on the remainder, and asks how well the hidden
hosts outrank never-observed species (Mann–Whitney AUC).

```{r holdout}
val_cal <- validate_host_prediction(interactions, phydist,
                                    parasite = caligus_sp, method = "pu",
                                    prop_holdout = 0.25, n_rep = 5,
                                    seed = 42)
val_lep <- validate_host_prediction(interactions, phydist,
                                    parasite = lepeo_sp, method = "pu",
                                    prop_holdout = 0.25, n_rep = 5,
                                    seed = 42)

data.frame(
  parasite    = c(caligus_sp, lepeo_sp),
  holdout_auc = c(mean(val_cal$auc, na.rm = TRUE),
                  mean(val_lep$auc, na.rm = TRUE)),
  sd          = c(sd(val_cal$auc, na.rm = TRUE),
                  sd(val_lep$auc, na.rm = TRUE))
)
```

## 6. Benchmark: Poulin & Mouillot's $S_{TD}$

As a model-free cross-check we compute Poulin & Mouillot's (2003)
phylogenetic distinctness of each parasite's realized host set, using
the same patristic distances.

```{r std}
spec <- host_specificity_index(interactions, phydist, group = "genus")
spec <- spec[!is.na(spec$std), ]
aggregate(std ~ group, data = spec, FUN = median)
```

Higher $S_{TD}$ = realized hosts more spread across the tree. In the
full comparative study $b^*$ and $S_{TD}$ correlate positively
(Spearman $\rho$ = 0.55–0.74) while capturing different constructs:
$b^*$ measures the *predicted suitable fraction* of the candidate
phylogeny at the model's optimal threshold, $S_{TD}$ only the spread
of the hosts already observed.

## Session info

```{r session}
sessionInfo()
```
